---
title: "Tidy data and the Tidyverse"
output:
  html_document:
    df_print: paged
---

Michael Dunn, 5th of April 2022

## Course overview and philosophy

- Linguists need a ‘good enough’ approach

  - Data carpentry, not cabinetmaking
  - Own your own analysis
  - Imagine what you want to do, not determined by the availability of tools
  - Understand what might be possible for a more expert collaborator
  
- Superpower: Linguists can understand computer languages as languages

- 🤨 Hmmm.....

## Revision

### Using RStudio

- Console pane
- Source pane
- Help/Files pane

### Learning strategies

- Like learning a natural language, you can't learn the language first and only *then* start speaking it. Find simple things to use R for and practice. Use it as your spreadsheet programme instead of excel. Use the console when you need a calculator.

- It can be hard to learn to ‘core’ of R, since it's not clear to a newcomer what that is. Start from what you know and expand outwards

- The internet is your friend. [Stack Overflow](https://stackoverflow.com) is a huge question-and-answer site for programming. You'll usually find that your questions have already been answered there.

- The **Help** tab in RStudio displays the output of the `help()` function. How do I use the `read_csv()` function? Use `help(read_csv)`! (in the console `?read_csv` is a shortcut to this) 

### Functions, Loading data

Functions are bundles of commands with a name. You can make your own functions (examples later), as well as use built-in functions and functions that come when your load a library.

At the top of most scripts you write you will want to load the tidyverse functions. You do this with the `library()` function. A function is a word with brackets at the end. If the function has arguments they are put inside the brackets. Thus, to load the tidyverse package you run:

```{r}
library(tidyverse) # load tidyverse if it's not already loaded
```
**Output**

  - If you don't have the tidyverse package installed the library function will give an error:
  *Error in library(tidyverse) : there is no package called ‘tidyverse’*
  - If it loads you'll see a report showing what's loaded (‘Attaching packages’ etc.)
  - If nothing happens it means you've loaded the library already (an example of ‘succeeding silently’)
  
- The `#` tells R not to read anything after the symbol. You can use it for explanatory comments, or to (perhaps temporarily) disable parts of your code

- The `read_csv` function can take as its argument the path to the file stored on your computer or as a url. These is also a `read_tsv` function which works exactly the same way, but it read *tab-delimited files* instead of *comma-delimited files*

- I've commented out the lines loading the data from urls below because I want to be able to work offline. Instead I'm using copies of the files I saved earlier to a directory called `data` next to the directory where I am working (the `..` means next directory up from where we currently are, so `../data` is the sister directory).

🤨 **TASK** Fix the following block of code so that it reads the data from the internet. Type e.g. `View(languages)` in the console to check the contents of the `languages` variable.

```{r}
# values <- read_csv("https://raw.githubusercontent.com/cldf-datasets/uratyp/v1.1/cldf/values.csv") 
# languages <- read_csv("https://raw.githubusercontent.com/cldf-datasets/uratyp/v1.1/cldf/languages.csv")
# parameters <- read_csv("https://raw.githubusercontent.com/cldf-datasets/uratyp/v1.1/cldf/parameters.csv")
values <- read_csv("../data/values.csv")
languages <- read_csv("../data/languages.csv")
parameters <- read_csv("../data/parameters.csv")
```

- Where am I working? Type `getwd()` (mnemonic ‘get working directory’) to find out. To set the working directory use `setwd("path/to/the/directory/where/I/want/to/work")` or (if you're using RStudio) choose an option from the menubar **Session** then **Set working directory ➤**

## Piping syntax

The output of one function can be the argument for another. Traditionally most computer language have done this by nesting functions. Thus, the `names` function gives your the names of the header row of a table. Since we've loaded the data from `languages.csv` as an object called `languages`, we could run:
```{r}
names(languages)
```

But we could also do this in one line, with naming the contents of the file, as follows:
```{r message=FALSE}
# the message=FALSE argument in the chunk definition stops RStudio from printing out 
# all the boring details of the data-loading process to the screen
names(read_csv("../data/languages.csv"))
```
Piping syntax is another way of expressing the same idea. The symbol `%>%` is called a pipe. The thing before the pipe is ‘sent’ to the thing after it as its first argument. The example above could be expressed with a pipe as:

```{r message=FALSE}
read_csv("../data/languages.csv") %>% names()
```

With the pipe operator the item on the left is ‘sent’ to the next function as its first argument. We could also write:

```{r}
languages %>% names()
```

Multiple items can be piped together, so we could even break the previous example up as:

```{r message=FALSE}
"../data/languages.csv" %>% read_csv() %>% names()
```
Note that "../data/languages.csv" is a character string, not a variable

## What is ‘tidy’ data?

- How does tidy data relate to graphical visualisation of data?

```{r echo=FALSE}
library(babynames) # you will have to install this if you want to replicate this code
babynames %>% 
  filter(name %in% c("Rick", "Morty", "Summer", "Beth", "Jerry")) %>%
  ggplot(aes(x=year, y=n, colour=name, linetype=sex)) +
  annotate("point", x=1960, y=11836, colour="red", size=4) +
  geom_line()
```
  * What do we know about the point under the red dot?
    * It's **olive green**, so it's the name "Jerry"
    * it's a **dotted line** a male name
    * the **x-axis value** shows that it is the year is 1960
    * the **y-axis value** count (n) is about 12000
  This is one row of data. One point on the plot == one row of the data
```{r echo=FALSE}
babynames %>% filter(name=="Jerry", year==1960, sex=="M")
```
  * the `aes()` command within `ggplot()` specifies which parameter of the data to map on to which graphical parameter: `x=year, y=n, colour=name, sex=linetype`

- What is ‘untidy’ data?
  
  * In short, untidy data has multiple datapoints per row. For example:
  
```{r echo=FALSE}
# Don't worry about what dark magic I am using here: I'm just taking some tidy 
# data and making it untidy for the purposes of illustration
babynames %>% 
  filter(name=="Jerry", year > 1950, year <= 1960) %>% 
  pivot_wider(id_cols=c("name", "sex"), names_from="year", values_from="n")
```
    You can't map this onto graphical parameters to make a plot like that above: how could you specify `year` in particular? 
    
  * Preview: untidy for one purpose might be tidy for another: . In tomorrow's lesson we will do a PCA analysis, which summarises many columns for a single language row
  * `pivot_longer()` and `pivot_wider()`

## Making graphs

- The `ggplot()` function starts planning the plot
- The `aes()` function is an argument of `ggplot()` which specifies which graphical dimensions will be mapped onto which dimensions of the (tidy!!!) data
- Other map layers and settings can be added to the `ggplot()` command using `+`. The various `geom_*()` functions take their parameters from ggplot/aes

```{r}
babynames %>% 
  filter(name %in% c("Rick", "Morty", "Summer", "Beth", "Jerry")) %>%
  ggplot(aes(x=year, y=n, colour=name, linetype=sex))  +
  geom_line()
```

- Histograms and density plots

Take the names from the year 2000 and make a histogram of how common they are:
```{r}
babynames %>% 
  filter(year==2000) %>% 
  ggplot(aes(x=n)) + geom_histogram()
```

```{r}
babynames %>% 
  filter(year==2000) %>% 
  ggplot(aes(x=n, fill=sex)) + 
  geom_histogram(position="dodge")
```

```{r}
babynames %>% 
  filter(year==2000) %>% 
  ggplot(aes(x=log10(n), fill=sex)) + geom_density(alpha=0.5)
```
- Bar charts
- Scatter plots
- Geographic scatterplots

## Tabular data

- excel data and why we hate it
- comma delimited and tab delimited files
- read_excel, read_tsv, read_csv
- loading uratyp

## Filtering rows and selecting columns

```{r}
languages
names(languages)
```
The `select` function is used to pick out columns
```{r}
# Specify the columns to include
languages %>%
  select(Name, Latitude, Longitude, Subfamily)
```
You can simultaneously rename columns during a select:
```{r}
# Specify the columns to include
languages %>%
  select(Language_Name=Name, Latitude, Longitude, Subgroup=Subfamily)
```

🤨 There is also a `rename` function that can rename columns without selecting (try this out yourself by renaming the `ISO639P3code` column to `ISO_Code`)

```{r}
#
```


Select can also be used negatively, with `-` signs:
```{r}
languages %>%
  select(-UT_Experts, -GB_Experts, -Source, -Macroarea)
```

🤨 Can you see why it's impossible to do a select with a mix of positive values and negative values?

The `filter` function removes/chooses rows:

```{r}
# Remember, double == is a test for equality, single = is used for arguments, and 
# as a synonym for <- (which sets a variable to a value)
languages %>% 
  filter(Subfamily == "Samoyed")
```

```{r}
# Not equals
finnougric <- languages %>%
  filter(Subfamily != "Samoyed")
finnougric
```

Other logical operators
```{r}
languages %>%
  filter(Latitude > 60)
```
- Filters can be piped together
- Multiple conditions can be included in a single filter function (by default joined by AND, but OR can be specified)
```{r}
languages %>%
  filter(Latitude > 60 & Latitude < 70) %>% # Logical AND is &
  filter(Subfamily == "Finnic" | Subfamily == "Saami") # Logical OR is |
```

## Adding columns

The function to add columns is called `mutate` (this isn't the most obvious name, but it's very useful and frequent so you'll get used to it):

```{r}
languages %>%
  mutate(Label=paste0(Name, "/", Subfamily)) %>%
  select(Label, Latitude, Longitude)
```

```{r}
languages %>%
  filter(Name == "Udmurt") %>%
  select(Name, Subfamily)
```

```{r}
babynames %>%
  filter(year==2000) %>% 
  ggplot(aes(x=n, fill=sex)) + geom_histogram(bins=5, position="dodge") + 
  scale_x_log10(labels=function(x) paste0(format(x, scientific=FALSE), "s")) 
```

