---
title: "Measuring Typological and Goegraphical Distances in R"
author: "Yingqi Jing"
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  # html_document:
  pdf_document:
     fig_crop: true
     fig_caption: true
     latex_engine: xelatex
     keep_tex: false
     citation_package: biblatex # important: will use autocite in latex
     toc: true
     toc_depth: 4
     number_section: true
link-citations: TRUE
---

\clearpage

```{r install and load the packages, include=F}
pkgs = c("knitr", "ggplot2", "dplyr", "tidyr", "tidyverse",
             "purrr", "janitor", "geosphere")
installed_packages = pkgs %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr) # gather & spread
library(tidyverse)
library(purrr)
library(janitor) # remove_constant
library(geosphere) # distHaversine
print("You are all good!")
```


R programming language
=====================

Conditional statement (`ifelse` & `case_when`)
-------------------------------------

```{r}
pi = 3.14
if(is.numeric(pi)){
  print("You have a numeric pi :)")
}else{
  print("You do not have a numeric pi :(")
}
```

```{r}
mylogic = TRUE
if(is.logical(mylogic)){
  print("You have a logical variable :)")
}else{
  print("You do not have a logical variable :(")
}
```


```{r}
lang = "Estonian"
if(is.character(lang) & lang == "Estonian"){
  print("It is a character variable and it is Estonian!")
}else{
  print("It is not Estonian!")
}
```

```{r}
lang = "Estonian"
if(is.character(lang) | lang == "Estonian"){
  print("It is a character or it is Estonian!")
}else{
  print("It is not Estonian!")
}
```


\clearpage

```{r}
lang_df = data.frame(id = 1:5,
          lang = c("Estonian", "Finnish", "English",
                   "Chinese", "Esperanto"))
lang_df
```

```{r if else statement}
lang_df %>%
  mutate(family = ifelse(lang == "Estonian", "Estonian", "not Estonian"))
```

We can generalize the `ifelse` into a case of multiple conditions, so that you have not only two options but multiple options.


```{r case when}
lang_df %>%
  mutate(family = case_when(lang %in% c("Estonian", "Finnish") ~ "Uralic",
                            lang %in% c("English") ~ "IE",
                            lang %in% c("Chinese") ~ "Sino-Tibetan",
                            TRUE ~ "unknown"))
```

\clearpage

For loop (`for` and `rowwise`)
-----------------------

```{r}
names = c("Yingqi", "John", "Emily", "David")
for(name in names){
  print(nchar(name))
}
```

Dplyr provides a convenient `rowwise` operation to work on each row of a data.frame. This can be viewed as an alternative to `for` loop in a data.frame. Note: `rowwise` groups your data by row (class: rowwise_df), and it is best to `ungroup` immediately.^[[https://medium.com/p/da3638b5f46c](https://medium.com/p/da3638b5f46c)]

To facilitate `rowwise` operation, you may need to use `c_across` to select specific columns. For example, if you want to calculate the number of characters of language names in a data.frame.

```{r}
mydf = data.frame(id = 1:4,
           names = c("Yingqi", "John", "Emily", "David"))
name_length = mydf %>%
  rowwise() %>%
  mutate(nchar_name = nchar(c_across(names))) %>%
  ungroup
name_length
```

\clearpage

Function
-------------


```{r}
myaverage = function(numvec){
  average = sum(numvec)/length(numvec)
  return(average)
}
```

```{r}
myvec = c(2, 3)
myaverage(myvec)
```


```{r calculate the average length of names}
name_length %>%
  pull(nchar_name) %>%
  myaverage()
```


\clearpage

Typological similarities between Uralic languages
========================================


Function
--------------

```{r}
typological_sim = function(data = ut_final, x, y){
  subdata = data %>%
    dplyr::select(-matches("subfamily|area", ignore.case = T)) %>%
    column_to_rownames(var = "Name")
  sim = sum(abs(as.vector(subdata[x, ]) == as.vector(subdata[y, ])))
  sim_p = sim/ncol(subdata)
  return(sim_p)
}
```


Data preprocessing
-------------------

```{r}
uratyp_df = read.csv("../data/values.csv")
lang_df = read.csv("../data/languages.csv")
ut_data = uratyp_df %>%
  inner_join(., lang_df, by = c("Language_ID" = "ID")) %>%
  dplyr::select(Name, Parameter_ID, Value, Subfamily) %>%
  filter(grepl("UT", Parameter_ID))
```

(1) Convert all data into binary (0, 1)

```{r}
ut_wide = ut_data %>%
  mutate(Value = case_when(
    Value == "0" ~ 0L,
    Value == "1" ~ 1L,
    TRUE ~ NA_integer_ # convert all "?" into NA
  )) %>%
  pivot_wider(., names_from = Parameter_ID, values_from = Value)
```


(2) Remove all columns with missing values

```{r}
ut_wide = ut_wide %>%
  select_if(function(x) !any(is.na(x)))
# alternatively, select_if(~ !any(is.na(.x)))
# select_if(~ sum(is.na(.x)) == 0)
# select(where(~ sum(is.na(.x)) == 0))
```

(3) Remove all constant columns

```{r}
ut_final = ut_wide %>%
  remove_constant(.)
  # select_if(~ length(unique(.x)) > 1)
head(ut_final)
```
(4) reshape the data into long format

```{r}
lang_sorted = ut_final %>%
  arrange(Subfamily) %>%
  pull(Name)
ut_final_long = ut_final %>%
  pivot_longer(., names_to = "feature", values_to = "value", -c("Name", "Subfamily")) %>%
  mutate(value = factor(value),
         Name = factor(Name, levels = lang_sorted))
head(ut_final_long)
```

(5) Visualize data via heatmap


```{r fig.width=14, fig.height=8, fig.cap="Overview of typological data in UT database"}
ut_final_long %>%
  ggplot(., aes(x = feature, y = Name, fill = value)) +
  geom_tile() +
  scale_fill_manual(values = alpha(c("blue", "red"), 0.65)) +
  theme(axis.text.x = element_text(angle = 90, size = 6, hjust = 0),
        axis.ticks = element_blank()) +
  labs(x = NULL, y = NULL)
```



\clearpage

Calculating typological similarities
---------------------------

```{r}
lgs = ut_final$Name
lgs_sim = expand.grid(lgs, lgs, stringsAsFactors = F) %>%
  rowwise() %>%
  mutate(similarity = typological_sim(ut_final, Var1, Var2))
```

```{r heatmap of mutual similarity between languages, fig.width=12, fig.height=12, fig.cap="Typological similarities between Uralic languages in UT data"}
lgs_sim_sorted = lgs_sim %>%
  mutate(Var1 = factor(Var1, levels = lang_sorted),
         Var2 = factor(Var2, levels = lang_sorted))
ggplot(lgs_sim_sorted, aes(Var1, Var2, fill = similarity)) +
  geom_tile() +
  geom_text(data = lgs_sim_sorted,
            mapping = aes(Var1, Var2,
                          label = round(similarity, digit = 1))) +
  scale_fill_continuous(type = "viridis") +
  labs(x = NULL, y = NULL) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 90, size = 9, hjust = 1),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  coord_fixed()
```
\clearpage

Exercises
-------------------

(1) Pls add the information of typological areas (*parameters.csv*) to UT dataset. Note: you can use the preprocessed UT data **ut_final_long**.

```{r}
feat_areas = read.csv("../Data/parameters.csv")
feat_areas = feat_areas %>%
  dplyr::select(ID, Area)
ut_areas = ut_final_long %>%
  inner_join(., feat_areas, by = c("feature" = "ID"))
```

(2) Pls calculate the typological similarities between languages across typological areas (phonology, morphology and syntax) in UT dataset, and plot them as heatmaps separately. Note: you can remove the lexicon features, and use `facet_wrap` function to create subpanels.


```{r}
ut_area_sim = ut_areas %>%
  filter(Area %in% c("Phonology", "Morphology", "Syntax")) %>%
  split(.$Area) %>%
  map_dfr(., ~{subdata = .x %>% spread(., key = feature, value = value)
               expand.grid(lgs, lgs, stringsAsFactors = F) %>%
                 rowwise() %>%
                 mutate(similarity = typological_sim(subdata, Var1, Var2)) %>%
                 mutate(Area = subdata$Area[1])})
```


```{r fig.width=18, fig.height=6, fig.cap="Typological similarities between Uralic languages across different areas in UT data"}
ut_area_sim_sorted = ut_area_sim %>%
  mutate(Var1 = factor(Var1, levels = lang_sorted),
         Var2 = factor(Var2, levels = lang_sorted),
         Area = factor(Area, levels = c("Phonology", "Morphology", "Syntax")))
ut_area_sim_sorted %>%
  ggplot(., aes(Var1, Var2, fill = similarity)) +
  geom_tile() +
  # geom_text(data = lgs_sim_sorted,
  #           mapping = aes(Var1, Var2,
  #                         label = round(similarity, digit = 1))) +
  scale_fill_continuous(type = "viridis") +
  facet_wrap(~Area) +
  # scale_x_discrete(position = "bottom") +
  labs(x = NULL, y = NULL) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 90, size = 9, hjust = 1),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  coord_fixed()
```


\clearpage

Geographical distances between languages
==============================

Function
-----------


```{r}
geographical_dist = function(data = lang_geo, x = lang1, y = lang2){
  lang1_location = data[x, ]
  lang2_location = data[y, ]
  return(distHaversine(lang1_location, lang2_location)/1000)
}
```


Measuring geographical distances
---------------------------------

```{r}
lang_geo = lang_df %>%
  dplyr::select(Name, Longitude, Latitude) %>%
  column_to_rownames(var = "Name")
```

```{r}
geo_dist = expand.grid(lgs, lgs, stringsAsFactors = F) %>%
  rowwise() %>%
  mutate(distance = geographical_dist(data = lang_geo,
                                      x = Var1,
                                      y = Var2)) %>%
  ungroup %>%
  mutate(dist_scaled = distance/max(distance))
geo_dist_sorted = geo_dist %>%
  mutate(Var1 = factor(Var1, levels = lang_sorted),
         Var2 = factor(Var2, levels = lang_sorted))
```


```{r geo distance, fig.width=12, fig.height=12, fig.cap="Geographical distances between Uralic languages"}
geo_dist_sorted %>%
  ggplot(., aes(Var1, Var2, fill = distance)) +
  geom_tile() +
  geom_text(data = geo_dist_sorted,
            mapping = aes(Var1, Var2,
                          label = round(dist_scaled, digit = 1))) +
  scale_fill_continuous(type = "viridis", direction = -1) +
  labs(x = NULL, y = NULL) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(angle = 90, size = 9, hjust = 1),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank()) +
  coord_fixed()
```
